Implementation and comparison of register allocation to translate \lstinline!LLVM--! to x86

### 2023-12-27 article.tex
Pre breaking ref

\usepackage{subcaption}
\usepackage[margin=1in]{geometry}
\usepackage{amsmath}
\usepackage[table]{xcolor}% http://ctan.org/pkg/xcolor
\usepackage{tikz}
\usetikzlibrary{angles}
\usetikzlibrary{arrows.meta}
\usetikzlibrary{automata}
\usetikzlibrary{calc}
\usetikzlibrary{chains}
\usetikzlibrary{decorations.pathreplacing}
\usetikzlibrary{positioning}
\usetikzlibrary{quotes}
\usepackage{graphicx} % Required for inserting images
\usepackage[T1]{fontenc}
\usepackage{beramono}
\usepackage{courier}
\usepackage{listings}
\usepackage{xpatch}
\usepackage{realboxes}
\usepackage{titlesec}
\usepackage{xcolor}
\usepackage{minted}
\usepackage{nicematrix}
\usepackage{collcell}
\usepackage{varwidth}

@@

Most CPUs are \textit{register machines}, in that they use a limited set of \textit{general-purpose registers} (GPRs) to store working values in combination with \textit{random access memory} (RAM) for mid-term, and other I/O peripherals for long-term storage. This can largely be attributed to performance, as register machines routinely outperform \textit{stack machines} \parencite{ShiYunhe2008VmsS} that are often used in VMs.
Although register machines generally also have a stack available, as being limited to mere bytes of storage is simply unfeasible for most large scale applications, using it compared to GPRs is several orders of magnitude slower.
Because of this, a crucial part of the backend stage for an optimizing compiler is assigning each variable  of the source program to a GPR in such a way that maximizes performance without sacrificing correctness.

@@

@@

@@

\subsection{Building a graph}

@@

With an input stream of instructions

@@

Another approach, which is a much more precise approximation, is a specific variant of the dataflow analysis as described in \textit{Modern Compiler Implementation in ML} (\cite{tiger}) and \textit{Compilers: Principles, Techniques, and Tools} (\cite{dragon}). In general, dataflow analysis is the process of finding the possible paths in which data may propagate through various branches of execution.  While several applications of this exist (like constant propagation, reaching defnitions, available expressions etc.), one that is immediately beneficial in the case of liveness analysis is one that traverses a CFG in the reverse order of execution (i.e.  \textit{backwards} flow), and extracts any variable that \textit{may} be used in execution (also referred to as \textit{backwards may} analysis). % FIXME: forklar hvorfor List.rev ikke udgør backwards flow

@@

let use (s : S.SS.t) (insn : Cfg.insn) =
  let op o s = match o with Ll.Id i -> S.SS.add i s | _ -> s in
  let po s o = op o s in
  match insn with
  | Insn (_, AllocaN (_, (_, o)))
  | Insn (_, Bitcast (_, o, _))
  | Insn (_, Load (_, o))
  | Insn (_, Ptrtoint (_, o, _))
  | Insn (_, Sext (_, o, _))
  | Insn (_, Trunc (_, o, _))
  | Insn (_, Zext (_, o, _)) ->
      op o s
  | Insn (_, Binop (_, _, l, r))
  | Insn (_, Icmp (_, _, l, r))
  | Insn (_, Store (_, l, r)) ->
      op l s |> op r
  | Insn (_, Call (_, _, args)) -> List.map snd args |> List.fold_left po s
  | Insn (_, Gep (_, bop, ops)) -> List.fold_left po (op bop s) ops
  | Insn (_, Select (c, (_, l), (_, r))) -> op c s |> op l |> op r
  | Insn (_, PhiNode (_, ops)) -> List.map fst ops |> List.fold_left po s
  | Term (Ret (_, Some o) | Cbr (o, _, _)) -> op o s
  | _ -> s
\end{minted}
Where \lstinline!S.SS! is a \lstinline!Set.S! module built over the \lstinline!symbol! type found in \lstinline!lib/symbol.ml!:
\begin{minted}[linenos,bgcolor=LightGray]{ocaml}
type symbol = string * int
(* ... *)
module SS = Set.Make (struct
  type t = symbol
  let compare (_, n1) (_, n2) = compare n1 n2
end)
type set = SS.t
\end{minted}
With the actual fixed-point iteration performed as follows:
\begin{minted}[linenos,bgcolor=LightGray]{ocaml}
let dataflow (insns : Cfg.insn list) (ids : Cfg.G.V.t array) (g : Cfg.G.t) =
  let insns = List.mapi (fun i v -> (i, v)) insns |> List.rev in
  let in_ = Array.init (List.length insns) (fun _ -> S.SS.empty) in
  let out = Array.init (List.length insns) (fun _ -> S.SS.empty) in
  let rec dataflow () =
    let flowout = (* ... *)
    let flowin = (* ... *)
    let flow changed insn = changed || flowout insn || flowin insn in
    if List.fold_left flow false insns then dataflow () else (in_, out)
  in
  dataflow ()
\end{minted}
The \texttt{flowin} function correponds to the \textit{live-in} equation \eqref{flowin} and is implemented as follows:
\begin{minted}[linenos,bgcolor=LightGray]{ocaml}
let flowin (i, insn) =
  let newin = S.SS.union (use insn) (S.SS.diff out.(i) (def insn)) in
  let changed = not (S.SS.equal newin in_.(i)) in
  if changed then in_.(i) <- newin;
  changed
\end{minted}
And the \texttt{flowout} function which correponds to the \textit{live-out} equation \eqref{flowout} implemented as follows:
\begin{minted}[linenos,bgcolor=LightGray]{ocaml}
let flowout (i, _) =
  let newout =
    let succ = Cfg.G.succ g ids.(i) in
    List.fold_left
      (fun s v -> S.SS.union s in_.(Cfg.G.V.label v))
      S.SS.empty succ
  in
  let changed = not (S.SS.equal newout out.(i)) in
  if changed then out.(i) <- newout;
  changed
\end{minted}

@@

Because of this, for any phi node of the form 
\[
  a = \Phi (b_1, ..., b_n)
\]
add edges to all \textit{live-out} variables not in \(B\):
\[
  \forall b_j \in (\mathit{out}[i] \setminus B), \mathit{add\_edge}(a, b_j) \text{ where } B = \{b_1, ..., b_n\}
\]
For any other instruction that defines a variable \(a\)
\[
  \forall b_j \in B, \mathit{add\_edge}(a, b_j).
\]
Although the Appel text (\cite{tiger}) describes interference of variables with concrete registers as well as overlapping variables, this isn't considered in this implementation.

@@

\section{Linear Scan}

%Write about Appel book.

@@

@@

\begin{NiceTabular}{>{\columncolor{gray!10}}ccccccccc}
\rowcolor{gray!10}n & clang & simple 12 & simple 2 & briggs 12 & briggs 2 & linear & greedy 12 & greedy 0 \\
20 & \cellcolor[rgb]{0.75,0.9990234375,0.75} 1.00& \cellcolor[rgb]{0.806640625,0.9990234375,0.75} 1.23& \cellcolor[rgb]{0.806640625,0.9990234375,0.75} 1.23& \cellcolor[rgb]{0.806640625,0.9990234375,0.75} 1.23& \cellcolor[rgb]{0.8056640625,0.9990234375,0.75} 1.23& \cellcolor[rgb]{0.8056640625,0.9990234375,0.75} 1.22& \cellcolor[rgb]{0.8056640625,0.9990234375,0.75} 1.22& \cellcolor[rgb]{0.9990234375,0.9912109375,0.75} 2.23\\
21 & \cellcolor[rgb]{0.75,0.9990234375,0.75} 1.00& \cellcolor[rgb]{0.8173828125,0.9990234375,0.75} 1.27& \cellcolor[rgb]{0.81640625,0.9990234375,0.75} 1.27& \cellcolor[rgb]{0.8173828125,0.9990234375,0.75} 1.27& \cellcolor[rgb]{0.8212890625,0.9990234375,0.75} 1.29& \cellcolor[rgb]{0.81640625,0.9990234375,0.75} 1.27& \cellcolor[rgb]{0.81640625,0.9990234375,0.75} 1.27& \cellcolor[rgb]{0.9990234375,0.9853515625,0.75} 2.44\\
22 & \cellcolor[rgb]{0.75,0.9990234375,0.75} 1.00& \cellcolor[rgb]{0.82421875,0.9990234375,0.75} 1.30& \cellcolor[rgb]{0.8232421875,0.9990234375,0.75} 1.29& \cellcolor[rgb]{0.8232421875,0.9990234375,0.75} 1.30& \cellcolor[rgb]{0.8212890625,0.9990234375,0.75} 1.29& \cellcolor[rgb]{0.8232421875,0.9990234375,0.75} 1.30& \cellcolor[rgb]{0.8232421875,0.9990234375,0.75} 1.29& \cellcolor[rgb]{0.9990234375,0.9794921875,0.75} 2.63\\
23 & \cellcolor[rgb]{0.75,0.9990234375,0.75} 1.00& \cellcolor[rgb]{0.8291015625,0.9990234375,0.75} 1.32& \cellcolor[rgb]{0.8291015625,0.9990234375,0.75} 1.32& \cellcolor[rgb]{0.8291015625,0.9990234375,0.75} 1.32& \cellcolor[rgb]{0.830078125,0.9990234375,0.75} 1.32& \cellcolor[rgb]{0.8310546875,0.9990234375,0.75} 1.33& \cellcolor[rgb]{0.8291015625,0.9990234375,0.75} 1.32& \cellcolor[rgb]{0.9990234375,0.9755859375,0.75} 2.75\\
24 & \cellcolor[rgb]{0.75,0.9990234375,0.75} 1.00& \cellcolor[rgb]{0.833984375,0.9990234375,0.75} 1.34& \cellcolor[rgb]{0.833984375,0.9990234375,0.75} 1.34& \cellcolor[rgb]{0.8349609375,0.9990234375,0.75} 1.34& \cellcolor[rgb]{0.8330078125,0.9990234375,0.75} 1.33& \cellcolor[rgb]{0.8330078125,0.9990234375,0.75} 1.34& \cellcolor[rgb]{0.8486328125,0.9990234375,0.75} 1.40& \cellcolor[rgb]{0.9990234375,0.97265625,0.75} 2.83\\
25 & \cellcolor[rgb]{0.75,0.9990234375,0.75} 1.00& \cellcolor[rgb]{0.8388671875,0.9990234375,0.75} 1.36& \cellcolor[rgb]{0.837890625,0.9990234375,0.75} 1.35& \cellcolor[rgb]{0.8369140625,0.9990234375,0.75} 1.35& \cellcolor[rgb]{0.8369140625,0.9990234375,0.75} 1.35& \cellcolor[rgb]{0.8359375,0.9990234375,0.75} 1.35& \cellcolor[rgb]{0.8359375,0.9990234375,0.75} 1.35& \cellcolor[rgb]{0.9990234375,0.970703125,0.75} 2.89\\
26 & \cellcolor[rgb]{0.75,0.9990234375,0.75} 1.00& \cellcolor[rgb]{0.837890625,0.9990234375,0.75} 1.36& \cellcolor[rgb]{0.8359375,0.9990234375,0.75} 1.35& \cellcolor[rgb]{0.8369140625,0.9990234375,0.75} 1.35& \cellcolor[rgb]{0.8369140625,0.9990234375,0.75} 1.35& \cellcolor[rgb]{0.837890625,0.9990234375,0.75} 1.36& \cellcolor[rgb]{0.8486328125,0.9990234375,0.75} 1.40& \cellcolor[rgb]{0.9990234375,0.9697265625,0.75} 2.92\\
27 & \cellcolor[rgb]{0.75,0.9990234375,0.75} 1.00& \cellcolor[rgb]{0.8388671875,0.9990234375,0.75} 1.36& \cellcolor[rgb]{0.837890625,0.9990234375,0.75} 1.35& \cellcolor[rgb]{0.83984375,0.9990234375,0.75} 1.36& \cellcolor[rgb]{0.8388671875,0.9990234375,0.75} 1.36& \cellcolor[rgb]{0.837890625,0.9990234375,0.75} 1.36& \cellcolor[rgb]{0.83984375,0.9990234375,0.75} 1.36& \cellcolor[rgb]{0.9990234375,0.96875,0.75} 2.96\\
28 & \cellcolor[rgb]{0.75,0.9990234375,0.75} 1.00& \cellcolor[rgb]{0.841796875,0.9990234375,0.75} 1.37& \cellcolor[rgb]{0.83984375,0.9990234375,0.75} 1.36& \cellcolor[rgb]{0.845703125,0.9990234375,0.75} 1.38& \cellcolor[rgb]{0.8427734375,0.9990234375,0.75} 1.37& \cellcolor[rgb]{0.8408203125,0.9990234375,0.75} 1.37& \cellcolor[rgb]{0.83984375,0.9990234375,0.75} 1.36& \cellcolor[rgb]{0.9990234375,0.96875,0.75} 2.97\\
29 & \cellcolor[rgb]{0.75,0.9990234375,0.75} 1.00& \cellcolor[rgb]{0.841796875,0.9990234375,0.75} 1.37& \cellcolor[rgb]{0.8408203125,0.9990234375,0.75} 1.37& \cellcolor[rgb]{0.841796875,0.9990234375,0.75} 1.37& \cellcolor[rgb]{0.83984375,0.9990234375,0.75} 1.36& \cellcolor[rgb]{0.8408203125,0.9990234375,0.75} 1.37& \cellcolor[rgb]{0.841796875,0.9990234375,0.75} 1.37& \cellcolor[rgb]{0.9990234375,0.9677734375,0.75} 2.98\\
30 & \cellcolor[rgb]{0.75,0.9990234375,0.75} 1.00& \cellcolor[rgb]{0.8349609375,0.9990234375,0.75} 1.35& \cellcolor[rgb]{0.8349609375,0.9990234375,0.75} 1.34& \cellcolor[rgb]{0.8359375,0.9990234375,0.75} 1.35& \cellcolor[rgb]{0.8330078125,0.9990234375,0.75} 1.34& \cellcolor[rgb]{0.8330078125,0.9990234375,0.75} 1.33& \cellcolor[rgb]{0.8349609375,0.9990234375,0.75} 1.34& \cellcolor[rgb]{0.9990234375,0.9697265625,0.75} 2.91\\
31 & \cellcolor[rgb]{0.75,0.9990234375,0.75} 1.00& \cellcolor[rgb]{0.83984375,0.9990234375,0.75} 1.36& \cellcolor[rgb]{0.8408203125,0.9990234375,0.75} 1.37& \cellcolor[rgb]{0.841796875,0.9990234375,0.75} 1.37& \cellcolor[rgb]{0.83984375,0.9990234375,0.75} 1.36& \cellcolor[rgb]{0.8408203125,0.9990234375,0.75} 1.37& \cellcolor[rgb]{0.83984375,0.9990234375,0.75} 1.36& \cellcolor[rgb]{0.9990234375,0.9677734375,0.75} 2.98\\
32 & \cellcolor[rgb]{0.75,0.9990234375,0.75} 1.00& \cellcolor[rgb]{0.8408203125,0.9990234375,0.75} 1.37& \cellcolor[rgb]{0.8408203125,0.9990234375,0.75} 1.37& \cellcolor[rgb]{0.8427734375,0.9990234375,0.75} 1.38& \cellcolor[rgb]{0.8408203125,0.9990234375,0.75} 1.37& \cellcolor[rgb]{0.841796875,0.9990234375,0.75} 1.37& \cellcolor[rgb]{0.8427734375,0.9990234375,0.75} 1.38& \cellcolor[rgb]{0.9990234375,0.9677734375,0.75} 2.99\\
33 & \cellcolor[rgb]{0.75,0.9990234375,0.75} 1.00& \cellcolor[rgb]{0.8427734375,0.9990234375,0.75} 1.37& \cellcolor[rgb]{0.841796875,0.9990234375,0.75} 1.37& \cellcolor[rgb]{0.8427734375,0.9990234375,0.75} 1.38& \cellcolor[rgb]{0.841796875,0.9990234375,0.75} 1.37& \cellcolor[rgb]{0.8408203125,0.9990234375,0.75} 1.37& \cellcolor[rgb]{0.841796875,0.9990234375,0.75} 1.37& \cellcolor[rgb]{0.9990234375,0.9677734375,0.75} 2.99\\
34 & \cellcolor[rgb]{0.75,0.9990234375,0.75} 1.00& \cellcolor[rgb]{0.841796875,0.9990234375,0.75} 1.37& \cellcolor[rgb]{0.841796875,0.9990234375,0.75} 1.37& \cellcolor[rgb]{0.841796875,0.9990234375,0.75} 1.37& \cellcolor[rgb]{0.841796875,0.9990234375,0.75} 1.37& \cellcolor[rgb]{0.8427734375,0.9990234375,0.75} 1.38& \cellcolor[rgb]{0.8408203125,0.9990234375,0.75} 1.37& \cellcolor[rgb]{0.9990234375,0.9677734375,0.75} 2.99\\
35 & \cellcolor[rgb]{0.75,0.9990234375,0.75} 1.00& \cellcolor[rgb]{0.841796875,0.9990234375,0.75} 1.37& \cellcolor[rgb]{0.841796875,0.9990234375,0.75} 1.37& \cellcolor[rgb]{0.8408203125,0.9990234375,0.75} 1.37& \cellcolor[rgb]{0.8427734375,0.9990234375,0.75} 1.37& \cellcolor[rgb]{0.841796875,0.9990234375,0.75} 1.37& \cellcolor[rgb]{0.8408203125,0.9990234375,0.75} 1.37& \cellcolor[rgb]{0.9990234375,0.9677734375,0.75} 2.99\\

>>>>

\usepackage{subcaption}
\usepackage[margin=1in]{geometry}
\usepackage{amsmath}
\usepackage[table]{xcolor}% http://ctan.org/pkg/xcolor
\usepackage{tikz}
\usetikzlibrary{angles}
\usetikzlibrary{arrows.meta}
\usetikzlibrary{automata}
\usetikzlibrary{calc}
\usetikzlibrary{chains}
\usetikzlibrary{decorations.pathreplacing}
\usetikzlibrary{positioning}
\usetikzlibrary{quotes}
\usepackage{graphicx} % Required for inserting images
\usepackage[T1]{fontenc}
\usepackage{beramono}
\usepackage{courier}
\usepackage{listings}
\usepackage{xpatch}
\usepackage{realboxes}
\usepackage{titlesec}
\usepackage{xcolor}
\usepackage{minted}
\usepackage{nicematrix}
\usepackage{collcell}
\usepackage{varwidth}
\usepackage{booktabs}

@@

Most CPUs are \textit{register machines}, in that they use a limited set of \textit{general-purpose registers} (GPRs) to store working values in combination with \textit{random access memory} (RAM) for mid-term, and other I/O peripherals for long-term storage. This can largely be attributed to performance, as register machines routinely outperform \textit{stack machines} \parencite{ShiYunhe2008VmsS} that are often used in VMs.
Although register machines generally also have a stack available, as being limited to mere bytes of storage is simply infeasible for most large scale applications, using it compared to GPRs is several orders of magnitudes slower.
Because of this, a crucial part of the backend stage for an optimizing compiler is assigning each variable  of the source program to a GPR in such a way that maximizes performance without sacrificing correctness.

@@

The implementation is written in OCaml for the most part, although there is some marginal use of Python to interact with the LLDB debugger.

@@

\section{Intermediate Representation}

Although compilers can feasibly translate from the source language to machine code of the target platform directly, and doing so could even be more efficient, doing so hinders the portability as machine code targeting architecture \(A\) isn't necessarily useful for targeting architecture \(B\) further down the line.

Instead of translating directly from a source language to target machine code, most compiler frontends emit IR which is intended as an abstraction over particular CPU implementations. % FIXME: er det korrekt? lol
It is by no means executable on either \(A\) or \(B\) but is more often than not much closer to the abstract level of a CPU than the source language originally fed to the frontend.  This helps portability immensely, as a frontend emitting an IR no longer needs to specialize to a particular architecture, instead it can target as many as the backend supports.
Likewise, a compiler backend will support any frontend provided that they emit correct IR.

Because of this, compilation of high-level programming languages will often go no further than emitting IR, then leave the rest for a subsequent backend toolchain. This naturally leads to the LLVM toolchain, which is by far the most used compiler backend. Most languages, if targeting native execution on a CPU, will have an LLVM implementation. This is true for C and C++ (clang and clang++ respectively), D (with the ldc compiler), Swift and Rust, to name a few.

Conversely, managed languages such as Java and C# typically don't depend on the LLVM toolchain directly, research projects that attempt to compile associated VM bytecode to a native platform using LLVM.

The modularization of the frontend and backend stages also help immensely.

\begin{figure}[!ht]
   \begin{minipage}{0.48\textwidth}
     \centering
     \begin{minted}[bgcolor=LightGray]{c}
int add(int a, int b) {
  int sum = a + b;
  return sum;
}
     \end{minted}
     \caption{Arithmetic function implemented in C}\label{fig:addc}
   \end{minipage}\hfill
   \begin{minipage}{0.48\textwidth}
     \centering
     \begin{minted}[bgcolor=LightGray]{llvm}
define i32 @add(i32 %a, i32 %b) {
  %sum = add i32 %a, %b
  ret i32 %sum
}
     \end{minted}
     \caption{Corresponding represented in LLVM}\label{fig:addll}
   \end{minipage}
   \begin{center}
   \begin{minipage}{0.48\textwidth}
     \centering
     \begin{minted}[bgcolor=LightGray]{gas}
  .globl _add
_add:
  movl	%edi, %eax
  addl	%esi, %eax
  retq
     \end{minted}
     \caption{Prettified result of \mintinline{bash}{clang -O0}}\label{fig:addx86}
   \end{minipage}
   \end{center}
\end{figure}

@@

\subsection{Building a CFG}

@@

Building a graph is relatively straight forward, with the input for every function declared parsed into a tuple type \mintinline{ocaml}{cfg}  of the form:
\begin{center}
\begin{minted}{ocaml}
type cfg = (lbl option * block) * (lbl * block) list
\end{minted}
\end{center}
which consists of an optionally named entry block in the first part and a list of trailing blocks that are always named in the second.

@@

Another approach, which is a more precise approximation, is a specific variant of the dataflow analysis as described in \textit{Modern Compiler Implementation in ML} (\cite{tiger}) and \textit{Compilers: Principles, Techniques, and Tools} (\cite{dragon}). In general, dataflow analysis is the process of finding the possible paths in which data may propagate through various branches of execution.  While several applications of this exist (like constant propagation, reaching defnitions, available expressions etc.), one that is immediately beneficial in the case of liveness analysis is one that traverses a CFG in the reverse order of execution (i.e.  \textit{backwards} flow), and extracts any variable that \textit{may} be used in execution (also referred to as \textit{backwards may} analysis). % FIXME: forklar hvorfor List.rev ikke udgør backwards flow

@@

let use (s : S.SS.t) (insn : Cfg.insn) =
  let op o s = match o with Ll.Id i -> S.SS.add i s | _ -> s in
  let po = Fun.flip op in
  match insn with
  | Insn (_, AllocaN (_, (_, o))) (* | Bitcast _ | ... | Zext _ *) ->
      op o s
  | Insn (_, Binop (_, _, l, r)) (* | Icmp _ | Store _ *) ->
      op l s |> op r
  | Insn (_, Call (_, _, args)) -> List.map snd args |> List.fold_left po s
  | Insn (_, Gep (_, bop, ops)) -> List.fold_left po (op bop s) ops
  | Insn (_, Select (c, (_, l), (_, r))) -> op c s |> op l |> op r
  | Insn (_, PhiNode (_, ops)) -> List.map fst ops |> List.fold_left po s
  | Term (Ret (_, Some o) | Cbr (o, _, _)) -> op o s
  | _ -> s
\end{minted}
Where \lstinline!S.SS! is a \lstinline!Set.S! module built over the \lstinline!symbol! type found in \lstinline!lib/symbol.ml!:
\begin{minted}[linenos,bgcolor=LightGray]{ocaml}
type symbol = string * int
(* ... *)
module SS = Set.Make (struct
  type t = symbol
  let compare (_, n1) (_, n2) = compare n1 n2
end)
type set = SS.t
\end{minted}
With the actual fixed-point iteration performed as follows:
\begin{minted}[linenos,bgcolor=LightGray]{ocaml}
let dataflow (insns : Cfg.insn list) (ids : Cfg.G.V.t array) (g : Cfg.G.t) =
  let insns = List.mapi (fun i v -> (i, v)) insns |> List.rev in
  let in_ = Array.init (List.length insns) (fun _ -> S.SS.empty) in
  let out = Array.init (List.length insns) (fun _ -> S.SS.empty) in
  let rec dataflow () =
    let flowout = (* ... *)
    let flowin = (* ... *)
    let flow changed insn = changed || flowout insn || flowin insn in
    if List.fold_left flow false insns then dataflow () else (in_, out)
  in
  dataflow ()
\end{minted}
The \texttt{flowin} function correponds to the \textit{live-in} equation \eqref{flowin} and is implemented as follows:
\begin{minted}[linenos,bgcolor=LightGray]{ocaml}
let flowin (i, insn) =
  let newin = S.SS.union (use insn) (S.SS.diff out.(i) (def insn)) in
  let changed = not (S.SS.equal newin in_.(i)) in
  if changed then in_.(i) <- newin;
  changed
\end{minted}
And the \texttt{flowout} function which correponds to the \textit{live-out} equation \eqref{flowout} implemented as follows:
\begin{minted}[linenos,bgcolor=LightGray]{ocaml}
let flowout (i, _) =
  let newout =
    let succ = Cfg.G.succ g ids.(i) in
    List.fold_left
      (fun s v -> S.SS.union s in_.(Cfg.G.V.label v))
      S.SS.empty succ
  in
  let changed = not (S.SS.equal newout out.(i)) in
  if changed then out.(i) <- newout;
  changed
\end{minted}

@@

Because of this, for any phi node of the form 
\[
  a = \Phi (b_1, ..., b_n)
\]
add edges to all \textit{live-out} variables not in \(B\):
\[
  \forall b_j \in \mathit{out}[i] \setminus \{b_1, ..., b_n\}, \text{add\_edge}(a, b_j)
\]
For any other instruction that defines a variable \(a\)
\[
  \forall b_j \in \{b_1, ..., b_n\}, \text{add\_edge}(a, b_j).
\]
Although the Appel text (\cite{tiger}) describes interference of variables with concrete registers as well as overlapping variables, this isn't considered in this implementation.

@@

\section{Linear Scan}
Another approach that does not depend on liveness analysis i the linear scan algorithm.
It is a simple and efficient approach that performs a linear pass over IR to produce a register assignment significantly faster than all of the steps involved with graph coloring.

The primary text is \textit{Linear Scan Register Allocation} (\cite{linear}), the execution of which also relies on prior liveness analysis.

% TODO: find ud af hvad binpacking er
%
%Write about Appel book.

@@

\begin{table}[!h]

@@

\begin{NiceTabular}{|c|cccccccc|}
\hline
n & clang & simple 12 & simple 2 & briggs 12 & briggs 2 & linear & greedy 12 & greedy 0 \\
\hline
8 & \cellcolor[rgb]{0.75,0.9990234375,0.75} 1.00000x& \cellcolor[rgb]{0.7529296875,0.9990234375,0.75} 1.01248x& \cellcolor[rgb]{0.75390625,0.9990234375,0.75} 1.01766x& \cellcolor[rgb]{0.75,0.9990234375,0.75} 1.00146x& \cellcolor[rgb]{0.7548828125,0.9990234375,0.75} 1.02261x& \cellcolor[rgb]{0.7509765625,0.9990234375,0.75} 1.00606x& \cellcolor[rgb]{0.7529296875,0.9990234375,0.75} 1.01436x& \cellcolor[rgb]{0.75,0.9990234375,0.75} 0.99702x\\
10 & \cellcolor[rgb]{0.75,0.9990234375,0.75} 1.00000x& \cellcolor[rgb]{0.75,0.9990234375,0.75} 0.99586x& \cellcolor[rgb]{0.7529296875,0.9990234375,0.75} 1.01369x& \cellcolor[rgb]{0.75,0.9990234375,0.75} 0.99508x& \cellcolor[rgb]{0.751953125,0.9990234375,0.75} 1.00861x& \cellcolor[rgb]{0.7548828125,0.9990234375,0.75} 1.01981x& \cellcolor[rgb]{0.751953125,0.9990234375,0.75} 1.00861x& \cellcolor[rgb]{0.755859375,0.9990234375,0.75} 1.02598x\\
12 & \cellcolor[rgb]{0.75,0.9990234375,0.75} 1.00000x& \cellcolor[rgb]{0.7548828125,0.9990234375,0.75} 1.02349x& \cellcolor[rgb]{0.76171875,0.9990234375,0.75} 1.04789x& \cellcolor[rgb]{0.759765625,0.9990234375,0.75} 1.04270x& \cellcolor[rgb]{0.7587890625,0.9990234375,0.75} 1.03716x& \cellcolor[rgb]{0.765625,0.9990234375,0.75} 1.06517x& \cellcolor[rgb]{0.763671875,0.9990234375,0.75} 1.05770x& \cellcolor[rgb]{0.77734375,0.9990234375,0.75} 1.11132x\\
14 & \cellcolor[rgb]{0.75,0.9990234375,0.75} 1.00000x& \cellcolor[rgb]{0.759765625,0.9990234375,0.75} 1.04054x& \cellcolor[rgb]{0.763671875,0.9990234375,0.75} 1.05851x& \cellcolor[rgb]{0.759765625,0.9990234375,0.75} 1.04050x& \cellcolor[rgb]{0.7626953125,0.9990234375,0.75} 1.05301x& \cellcolor[rgb]{0.7568359375,0.9990234375,0.75} 1.02863x& \cellcolor[rgb]{0.7607421875,0.9990234375,0.75} 1.04599x& \cellcolor[rgb]{0.7900390625,0.9990234375,0.75} 1.16166x\\
16 & \cellcolor[rgb]{0.75,0.9990234375,0.75} 1.00000x& \cellcolor[rgb]{0.771484375,0.9990234375,0.75} 1.08630x& \cellcolor[rgb]{0.7724609375,0.9990234375,0.75} 1.09112x& \cellcolor[rgb]{0.7666015625,0.9990234375,0.75} 1.06935x& \cellcolor[rgb]{0.7724609375,0.9990234375,0.75} 1.09077x& \cellcolor[rgb]{0.771484375,0.9990234375,0.75} 1.08669x& \cellcolor[rgb]{0.765625,0.9990234375,0.75} 1.06631x& \cellcolor[rgb]{0.8525390625,0.9990234375,0.75} 1.41519x\\
18 & \cellcolor[rgb]{0.75,0.9990234375,0.75} 1.00000x& \cellcolor[rgb]{0.7919921875,0.9990234375,0.75} 1.17172x& \cellcolor[rgb]{0.791015625,0.9990234375,0.75} 1.16684x& \cellcolor[rgb]{0.7880859375,0.9990234375,0.75} 1.15447x& \cellcolor[rgb]{0.7900390625,0.9990234375,0.75} 1.16224x& \cellcolor[rgb]{0.7919921875,0.9990234375,0.75} 1.17110x& \cellcolor[rgb]{0.7900390625,0.9990234375,0.75} 1.16144x& \cellcolor[rgb]{0.9560546875,0.9990234375,0.75} 1.83133x\\
20 & \cellcolor[rgb]{0.75,0.9990234375,0.75} 1.00000x& \cellcolor[rgb]{0.806640625,0.9990234375,0.75} 1.23037x& \cellcolor[rgb]{0.80859375,0.9990234375,0.75} 1.23533x& \cellcolor[rgb]{0.8076171875,0.9990234375,0.75} 1.23433x& \cellcolor[rgb]{0.8095703125,0.9990234375,0.75} 1.23986x& \cellcolor[rgb]{0.806640625,0.9990234375,0.75} 1.23120x& \cellcolor[rgb]{0.8046875,0.9990234375,0.75} 1.22283x& \cellcolor[rgb]{0.9990234375,0.990234375,0.75} 2.26430x\\
22 & \cellcolor[rgb]{0.75,0.9990234375,0.75} 1.00000x& \cellcolor[rgb]{0.826171875,0.9990234375,0.75} 1.30843x& \cellcolor[rgb]{0.8251953125,0.9990234375,0.75} 1.30498x& \cellcolor[rgb]{0.826171875,0.9990234375,0.75} 1.30814x& \cellcolor[rgb]{0.826171875,0.9990234375,0.75} 1.30594x& \cellcolor[rgb]{0.8251953125,0.9990234375,0.75} 1.30564x& \cellcolor[rgb]{0.82421875,0.9990234375,0.75} 1.29918x& \cellcolor[rgb]{0.9990234375,0.978515625,0.75} 2.64427x\\
24 & \cellcolor[rgb]{0.75,0.9990234375,0.75} 1.00000x& \cellcolor[rgb]{0.8408203125,0.9990234375,0.75} 1.36606x& \cellcolor[rgb]{0.83984375,0.9990234375,0.75} 1.36299x& \cellcolor[rgb]{0.8408203125,0.9990234375,0.75} 1.36538x& \cellcolor[rgb]{0.83984375,0.9990234375,0.75} 1.36151x& \cellcolor[rgb]{0.83984375,0.9990234375,0.75} 1.36449x& \cellcolor[rgb]{0.83984375,0.9990234375,0.75} 1.36204x& \cellcolor[rgb]{0.9990234375,0.970703125,0.75} 2.89705x\\
26 & \cellcolor[rgb]{0.75,0.9990234375,0.75} 1.00000x& \cellcolor[rgb]{0.8369140625,0.9990234375,0.75} 1.35207x& \cellcolor[rgb]{0.8369140625,0.9990234375,0.75} 1.35059x& \cellcolor[rgb]{0.837890625,0.9990234375,0.75} 1.35397x& \cellcolor[rgb]{0.8369140625,0.9990234375,0.75} 1.34954x& \cellcolor[rgb]{0.8369140625,0.9990234375,0.75} 1.35255x& \cellcolor[rgb]{0.8369140625,0.9990234375,0.75} 1.34932x& \cellcolor[rgb]{0.9990234375,0.9697265625,0.75} 2.92968x\\
28 & \cellcolor[rgb]{0.75,0.9990234375,0.75} 1.00000x& \cellcolor[rgb]{0.8388671875,0.9990234375,0.75} 1.35918x& \cellcolor[rgb]{0.837890625,0.9990234375,0.75} 1.35626x& \cellcolor[rgb]{0.8388671875,0.9990234375,0.75} 1.36006x& \cellcolor[rgb]{0.837890625,0.9990234375,0.75} 1.35674x& \cellcolor[rgb]{0.8388671875,0.9990234375,0.75} 1.35812x& \cellcolor[rgb]{0.837890625,0.9990234375,0.75} 1.35462x& \cellcolor[rgb]{0.9990234375,0.96875,0.75} 2.96506x\\
30 & \cellcolor[rgb]{0.75,0.9990234375,0.75} 1.00000x& \cellcolor[rgb]{0.8515625,0.9990234375,0.75} 1.41137x& \cellcolor[rgb]{0.8505859375,0.9990234375,0.75} 1.40745x& \cellcolor[rgb]{0.8515625,0.9990234375,0.75} 1.41125x& \cellcolor[rgb]{0.8505859375,0.9990234375,0.75} 1.40768x& \cellcolor[rgb]{0.8515625,0.9990234375,0.75} 1.40935x& \cellcolor[rgb]{0.8505859375,0.9990234375,0.75} 1.40589x& \cellcolor[rgb]{0.9990234375,0.96484375,0.75} 3.08506x\\
32 & \cellcolor[rgb]{0.75,0.9990234375,0.75} 1.00000x& \cellcolor[rgb]{0.853515625,0.9990234375,0.75} 1.41590x& \cellcolor[rgb]{0.8525390625,0.9990234375,0.75} 1.41255x& \cellcolor[rgb]{0.853515625,0.9990234375,0.75} 1.41652x& \cellcolor[rgb]{0.8525390625,0.9990234375,0.75} 1.41300x& \cellcolor[rgb]{0.8525390625,0.9990234375,0.75} 1.41399x& \cellcolor[rgb]{0.8515625,0.9990234375,0.75} 1.41115x& \cellcolor[rgb]{0.9990234375,0.9638671875,0.75} 3.09890x\\
\hline
